Clearance Project - 4 - PhysChem Clustering

### 2025-03-11 data-raw/ClearanceProject-4-PhysChemClustering.Rmd
Renamed and initialized 4 analysis documents for clearance project.

---
title: "Restrictive Clearance Comparison"
author: "Gilberto Padilla Mercado"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

This vignette describes comparisons between invivoPKfit predicted and those from httk.

```{r library_loading}
today <- format(Sys.Date(), "%d%b%Y")
devtools::load_all()
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(ctxR)
library(ggvenn)
library(ggalluvial)
library(flextable)
library(reticulate)
library(patchwork)
source("data-raw/RC_comparison_bkgd.R") # httk & dawson_2021 loaded here
```

We first want to see how many chemicals are in either _httk_ or _invivoPKfit_

```{r venn_diagrams}
member_list <- list(
  httk_3comp2 = unique(get_cheminfo(info = "DTXSID",
                                species = "human",
                                model = "3compartment2",
                                median.only = TRUE,
                                physchem.exclude = TRUE,
                                suppress.messages = TRUE)) %>% 
    suppressWarnings(),
  cvt = cvt %>% 
    filter(species %in% c("human", "rat"),
           conc_medium_normalized == "plasma") %>%
    pull(analyte_dtxsid)
)

ggvenn(member_list, fill_color = c("white", "white"))

ggsave(filename = paste0(Sys.getenv("FIG_DIR"),
                         "RestrictiveClearance_Analysis/",
                         "CURRENT_", "httk_vs_CvT_VennDiagram_human.png"),
       bg = "white",
       height = 3, width = 5
)

target_chems <- intersect(member_list$cvt, member_list$httk_3comp2)

filter_targets <- function(.data) {
  stopifnot("Chemical" %in% names(.data))
  dplyr::filter(.data = .data, Chemical %in% target_chems)
}
```

I select which parameterization (based on species) that I will use.

```{r selecting_chemical_params}
chosen_list <- rat_pars_human
# The following are derived from the above selected list (see environment)
this_chem_v <- unique(chosen_list$param_df$Chemical)
this_param_rest_l <- chosen_list$named_params_restrictive
this_param_nrest_l <- chosen_list$named_params_nonrestrictive
this_param_df <- chosen_list$param_df
this_httk_clearance_df <- chosen_list$clearance_df
this_species_v <- c("rat", "human")
# Testing to make sure there are some chemicals that have different clearances
count(this_httk_clearance_df, restrictive_clearance == nonrestrictive_clearance)
# The majority (140) do!
```

Here I do a quick fit of the CvT data only for those that have plasma measurements.

```{r quick-fit}

# Only use plasma for simplicity
this_pk <- pk(cvt %>% 
                dplyr::filter(conc_medium_normalized == "plasma")
              ) +
  settings_optimx(method = "bobyqa") +
  facet_data(vars(Chemical, Species)) +
  settings_preprocess(
    keep_data_original = TRUE,
    suppress.messages = FALSE) +
  scale_conc(dose_norm = TRUE, log10_trans = TRUE) +
  stat_error_model(error_group = vars(Chemical, Species))

this_pk <- do_prefit(this_pk)
this_pk <- do_fit(this_pk, n_cores = 14)
```

```{r load-other}
# Here I want to load all other data and setup other data objects
# Kd/binding affinity data
load(paste0(Sys.getenv("OD_DIR"),
            "Profile/Documents/New-HTTK-raw-2019-05-06.RData")
)

# Chemical info and properties (averaging the latter by chemical and experimental/predicted)
target_chem_info <- get_chemical_details_batch(DTXSID = target_chems,
                                               rate_limit = 1L)
target_chem_properties <- get_chem_info_batch(DTXSID = target_chems, 
                                              rate_limit = 1L) %>%
  filter(name %in% c("LogKow: Octanol-Water",
                     "pKa Acidic Apparent",
                     "pKa Basic Apparent")) %>% 
  group_by(dtxsid, propType, propertyId) %>%
  summarize(Mean = mean(value, na.rm = TRUE),
            StDev = sd(value, na.rm = TRUE)) %>%
  ungroup() %>%
  distinct()

# Bring together all chemical properties
target_chem_pw_predicted <- target_chem_properties %>%
  filter(propType == "predicted") %>%
  pivot_wider(names_from = propertyId,
              names_prefix = "pr_",
              id_cols = dtxsid,
              values_from = c("Mean", "StDev"))
target_chem_pw_experimental <- target_chem_properties %>%
  filter(propType == "experimental") %>%
  pivot_wider(names_from = propertyId,
              names_prefix = "ex_",
              values_from = c("Mean", "StDev"),
              id_cols = dtxsid)

target_chem_pw_full <- full_join(
  target_chem_pw_experimental,
  target_chem_pw_predicted
) %>%
  janitor::clean_names() %>%
  mutate(dtxsid,
         type_logkow = ifelse(is.na(mean_ex_logkow_octanol_water),
                              NA, "experimental"),
         logkow = ifelse(is.na(mean_ex_logkow_octanol_water),
                         mean_pr_logkow_octanol_water,
                         mean_ex_logkow_octanol_water),
         type_apKa = ifelse(is.na(mean_ex_pka_acidic_apparent),
                            NA, "experimental"),
         apKa = ifelse(is.na(mean_ex_pka_acidic_apparent),
                       mean_pr_pka_acidic_apparent,
                       mean_ex_pka_acidic_apparent),
         type_bpKa = ifelse(is.na(mean_ex_pka_basic_apparent),
                            NA, "experimental"),
         bpKa = ifelse(is.na(mean_ex_pka_basic_apparent),
                       mean_pr_pka_basic_apparent,
                       mean_ex_pka_basic_apparent),
         .keep = "unused") %>%
  mutate(type_logkow = ifelse(!is.na(logkow) & is.na(type_logkow),
                              "predicted",
                              ifelse(type_logkow == "experimental",
                                     "experimental", NA)),
         type_apKa = ifelse(!is.na(apKa) & is.na(type_apKa),
                            "predicted",
                            ifelse(type_apKa == "experimental",
                                     "experimental", NA)),
         type_bpKa = ifelse(!is.na(bpKa) & is.na(type_bpKa),
                            "predicted",
                            ifelse(type_bpKa == "experimental",
                                     "experimental", NA))) %>%
  mutate(logkow_sd = ifelse(type_logkow == "experimental",
                            st_dev_ex_logkow_octanol_water,
                            st_dev_pr_logkow_octanol_water),
         apKa_sd = ifelse(type_apKa == "experimental",
                          st_dev_ex_pka_acidic_apparent,
                          st_dev_pr_pka_acidic_apparent),
         bpKa_sd = ifelse(type_bpKa == "experimental",
                          st_dev_ex_pka_basic_apparent,
                          st_dev_pr_pka_basic_apparent)) %>%
  select(-starts_with("st_dev")) %>%
  distinct()
# Remove the parts of this final dataset
rm(target_chem_pw_experimental, target_chem_pw_predicted)

# Predictions from httk(code retained but not run after 2.20.2025)
restrictive_preds <- get_httk_preds(
  parameters = this_param_rest_l,
  pk_obj = this_pk,
  species = c("rat", "human")
) %>% rename_with(
  .fn = ~paste0(.x, "_restrictive"),
  .cols = starts_with("httk_")
)

nonrestrictive_preds <- get_httk_preds(
  parameters = this_param_nrest_l,
  pk_obj = this_pk,
  species = c("rat", "human")
) %>% rename_with(
  .fn = ~paste0(.x, "_nonrestrictive"),
  .cols = starts_with("httk_")
)

names(restrictive_preds)

all_httk_preds <- inner_join(restrictive_preds, nonrestrictive_preds) %>%
  filter_targets()
rm(restrictive_preds, nonrestrictive_preds)
# 
readr::write_csv(all_httk_preds,
                 file = paste0(Sys.getenv("FIG_DIR"),
                               "RestrictiveClearance_Analysis/",
                               "CURRENT_all_httk_preds.csv"))

all_httk_preds <- readr::read_csv(file = paste0(Sys.getenv("FIG_DIR"), 
                                                "RestrictiveClearance_Analysis/",
                                                "CURRENT_all_httk_preds.csv"))

this_nonflat_winmodel <- get_winning_model(this_pk) %>%
  filter(model %in% c("model_1comp", "model_2comp")) %>%
  filter_targets()

# Get the winning model (including flats) and only include chemicals in target_chems
this_winmodel <- get_winning_model(this_pk) %>%
    filter_targets()

winning_preds <- predict(this_pk) %>%
  semi_join(this_nonflat_winmodel) %>% # The targets already filtered here
  distinct(Chemical, Species, Chemical_Name,
           model, method, Dose, Media, Route, Reference, pLOQ, Time,
           N_Subjects, Conc, Conc_est)

# TKstats for 1-/2-compartment model
this_tkstats <- eval_tkstats(this_pk,
                             model = c("model_1comp", "model_2comp"),
                             finite_only = TRUE,
                             dose_norm = TRUE
) %>%
  filter_targets()
colnames(this_tkstats) <- gsub("/", "_", colnames(this_tkstats))
this_tkstats <- this_tkstats %>%
  inner_join(this_winmodel)

# Fraction unbound data for target chemicals
fup_df <- this_param_df %>%
  distinct(Chemical, Species, forced_human_values,
           Funbound.plasma, Funbound.plasma.adjustment, 
           Funbound.plasma.dist) %>%
  mutate(
    Funbound.plasma.median = as.numeric(str_extract(
      Funbound.plasma.dist, pattern = "^\\d+\\.?\\d+e?-?\\d+(?=,)"
    )),
    Funbound.plasma.lower95 = as.numeric(str_extract(
      Funbound.plasma.dist, pattern = "(?<=,)\\d+\\.?\\d+e?-?\\d+(?=,)"
    )),
    Funbound.plasma.upper95 = as.numeric(str_extract(
      Funbound.plasma.dist, pattern = "(?<=,)\\d+\\.?\\d+e?-?\\d$"
    ))
    )

human_chems <- httk::get_cheminfo(info = "DTXSID",
                                species = "human",
                                model = "3compartment2",
                                median.only = TRUE,
                                physchem.exclude = TRUE,
                                suppress.messages = TRUE)
human_chems <- intersect(human_chems, new.raw.httk.data$DTXSID)

all_hum_rat_fups <- chem.physical_and_invitro.data %>%
  filter(DTXSID %in% human_chems) %>%
  select(DTXSID, Human.Funbound.plasma, Rat.Funbound.plasma) %>%
  mutate(Rat.Funbound.plasma = as.numeric(str_extract(Rat.Funbound.plasma,
                                                      "^([^,]+)")),
         Human.Funbound.plasma = as.numeric(str_extract(Human.Funbound.plasma,
                                                        "^([^,]+)"))
         )

# Get restrictive and non-restrictive parameters
human_chems_fup <- data.frame(
  Chemical = human_chems,
  fup_restrictive = chem.physical_and_invitro.data$Human.Funbound.plasma[],
                              fup_nonrestrictive = params_nr)
```

Relationships in the mathematical PK models:

- When there is only oral data available, `Fgutabs_Vdist` and `CLtot_Fgutabs`
will be calculated instead of the numerator alone.  
- $C_{ss} \cdot CL_{tot} = C_{H,u}\cdot CL{int}$ (Bennet & Sodhi 2024)  
- [Add More Here]

The first thing I need to do is get the chemicals included in both the CvTdb subset
and the httk empirical data for pbpk.

Next I need to get the clearances using the `calc_total_clearance(dtxsid, restrictive.clearance)`
function from httk. Also need the clearances from invivoPKfit's winning models.

Note: There is about 50 chemicals with rat data for both, but 3X that when
we allow substitution of human numbers. However, this does not necessarily mean
that we will get appropriate fits with the human numbers.
The aforementioned set of rat data, for example, 33 of 51 chemicals find
appropriate fits with invivoPKfit + httk.

```{r load_Kd}
new.raw.httk.data %>%
  distinct(DTXSID, Affinity.Kd.Med, Affinity.Kd.Low, Affinity.Kd.High,
           Fit) %>%
  na.omit() %>%
  mutate(Target_Chemical = (DTXSID %in% target_chems)) %>%
ggplot(
       aes(
         x = forcats::fct_reorder(DTXSID, Affinity.Kd.Low)
         )) +
  geom_linerange(aes(
    ymin  = Affinity.Kd.Low,
    ymax = Affinity.Kd.High,
    color = Target_Chemical
  ), alpha = 0.4) +
  geom_point(aes(
    y = Affinity.Kd.Med, color = Target_Chemical
  ), size = 0.75) +
  # facet_grid(cols = vars(Fit), space = "free_x", scales = "free_x") +
  scale_y_log10(labels = scales::label_log(digits = 1)) +
  scale_color_manual(values = c("black", "green4")) +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid = element_blank(),
        plot.title = element_text(hjust = 0.5, face = "bold"),
        legend.position = "bottom") +
  labs(title = "Kd estimates across chemicals", x = NULL)

ggsave(filename = paste0(Sys.getenv("FIG_DIR"),
                         "RestrictiveClearance_Analysis/",
                         "CURRENT_", "Kd_estimates_initialsummary.png"),
       bg = "white",
       height = 4, width = 9
)

kd_v_fup_plot <- new.raw.httk.data %>%
  inner_join(all_hum_rat_fups) %>%
  pivot_longer(cols = Human.Funbound.plasma:Rat.Funbound.plasma,
               names_to = "Species", values_to = "httk_fup_med") %>%
  ggplot(aes(x = Affinity.Kd.Med, y = httk_fup_med, color = Species)) +
  geom_point(alpha = 0.5) +
  scale_color_manual(values = c("green4", "purple2"),
                     labels = c("Human", "Rat")) +
  scale_x_log10(labels = scales::label_log(digits = 2)) +
  labs(x = "Binding affinity ratio", y = "Fraction unbound\nin plasma") +
  theme_bw() +
  theme(legend.position = "top",
        legend.text = element_text(size = 10),
        legend.title = element_blank(),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))

ggsave(filename = paste0(Sys.getenv("FIG_DIR"),
                         "RestrictiveClearance_Analysis/",
                         "CURRENT_", "Kd_estimates_vs_fup_multispecies.png"),
       bg = "white",
       height = 3, width = 5
)

@@

target_chem_properties %>%
  ggplot(aes(x = propType, y = Mean)) +
  geom_violin() +
  geom_line(aes(group = dtxsid), color = "grey10", alpha = 0.1) +
  geom_hline(yintercept = 5, linetype = "longdash") +
  facet_grid(scales = "free_x",
             rows = vars(propertyId)) +
  scale_y_log10() +
  theme_bw() +
  theme(axis.text.x = element_text(hjust = 1, angle = 45))

target_chem_properties %>%
  count(source)

target_chem_info %>%
  ggplot(aes(x = monoisotopicMass)) +
  geom_histogram(binwidth = 50, fill = "grey5",
                 boundary = 0) +
  geom_vline(xintercept = 500, linetype = "longdash") +
  scale_x_continuous(limits = c(0, 1000),
                     breaks = seq(0, 1000, 100)) +
  theme_bw()

@@

```{r reticulate-prep}
smiles_list <- target_chem_info %>% pull(smiles) %>% unique()
mykit <- import("rdkit")
```

@@

```{r lipinksi_alluvial}

target_chem_summary <- target_chem_info %>%
  distinct(dtxsid, monoisotopicMass, smiles) %>%
  right_join(data.frame(smiles = smiles_list)) %>%
  bind_cols(data.frame(H_donors = py_to_r(py$py_h_donors),
                       H_acceptors = py_to_r(py$py_h_acceptors),
                       TPSA = py_to_r(py$py_tpsa),
                       Rot_Bonds = py_to_r(py$py_rotbonds))
            ) %>%
  left_join(
    target_chem_pw_full
  ) %>%
  rowwise() %>%
  mutate(logD_7.4 = httk::calc_dow(dtxsid = dtxsid, pH = 7.4))

target_chem_lipinski <- target_chem_summary %>%
  select(-c(smiles)) %>%
  group_by(dtxsid) %>%
  mutate("Mass <= 500 Da" = (monoisotopicMass <= 500),
          "H-donors < 5" = (H_donors < 5),
          "H-acceptors < 10" = (H_acceptors < 10),
          "logP < 5" = (logkow < 5),
          .keep = "none") %>%
  ungroup() %>%
  mutate(Lipinski = rowSums(dplyr::select(., where(is.logical))) >= 3)

lipinski_plot <- target_chem_lipinski %>%
  group_by(Lipinski) %>%
  count(`Mass <= 500 Da`, `H-donors < 5`, `H-acceptors < 10`, `logP < 5`,
        name = "Count") %>%
  ggplot(aes(
    axis3 = `Mass <= 500 Da`,
    axis4 = `logP < 5`,
    axis2 = `H-donors < 5`,
    axis1 = `H-acceptors < 10`,
    y = Count
  )) +
  geom_alluvium(aes(fill = Lipinski)) +
  geom_stratum() +
  scale_x_discrete(limits = c( 
    "H-acceptors < 10",
    "H-donors < 5",
    "<= 500 Da",
    "logP < 5"
  ),
  position = "top",
  expand = c(.2, .05)) +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)),
            size = 2) +
  labs(title = "Target chemicals passing Lipinksi's Rules of 5",
       fill = "Passes Lipinksi's Rules?") +
  scale_fill_manual(values = c("grey20", "green4")) +
  theme_minimal() +
  theme(axis.text.x = element_text(face = "bold", color = "black", size = 10),
        panel.grid = element_blank(),
        plot.title = element_text(hjust = 0.5),
        legend.position = "bottom")

lipinski_plot
ggsave(filename = paste0(Sys.getenv("FIG_DIR"),
                         "RestrictiveClearance_Analysis/",
                         "CURRENT_", "Alluvium_Lipinski.png"),
       bg = "white",
       height = 4, width = 5
)

@@

We also want to assess whether the "winning" _invivoPKfit_ fits are better than
the chosen _httk_ fit. This is a bit of a pre-requisite to our analysis of the
clearances and whether they are more or less reasonable in their reflection of
the reality. The metric used here will be RMSLE.

```{r httk_pr_rmsle}

# RMSLE formula
# LE = log1p(Conc_est) - log1p(Conc)
# SLE = LE^2
# RMSLE = sqrt(mean(SLE, na.rm = TRUE))

combined_preds <- all_httk_preds %>%
  mutate(Reference = as.character(Reference)) %>%
  left_join(winning_preds) %>%
  filter(!is.na(model))

combined_RMSLE <- combined_preds %>%
  ungroup() %>%
  mutate(
    restrictive_SLE = (log1p(httk_Preds_restrictive) - log1p(Conc))^2,
    nonrestrictive_SLE = (log1p(httk_Preds_nonrestrictive) - log1p(Conc))^2,
    invivoPK_SLE =  (log1p(Conc_est) - log1p(Conc))^2 
  ) %>%
  group_by(Chemical, Species, Route) %>%
  summarize(RMSLE_restrictive = sqrt(mean(restrictive_SLE, na.rm = TRUE)),
            RMSLE_nonrestrictive = sqrt(mean(nonrestrictive_SLE, na.rm = TRUE)),
            RMSLE_invivoPKfit = sqrt(mean(invivoPK_SLE, na.rm = TRUE)))

combined_RMSLE %>%
  ungroup() %>%
  pivot_longer(cols = starts_with("RMSLE_"),
               names_to = "Source",
               values_to = "RMSLE") %>%
  mutate(Source = factor(Source,
                         levels = c("RMSLE_nonrestrictive",
                                    "RMSLE_invivoPKfit",
                                    "RMSLE_restrictive")
  )) %>%
  ggplot(aes(x = Source, y = RMSLE, group = Chemical)) +
  geom_point(alpha = 0.5) +
  geom_line(alpha = 0.5) +
  theme_bw()

combined_RMSLE %>%
  ungroup() %>%
  mutate(comparison_group = case_when(
    between(RMSLE_invivoPKfit, RMSLE_nonrestrictive, RMSLE_restrictive) |
     between(RMSLE_invivoPKfit, RMSLE_restrictive, RMSLE_nonrestrictive) ~ "Between",
    RMSLE_invivoPKfit < RMSLE_nonrestrictive &
      RMSLE_invivoPKfit < RMSLE_restrictive ~ "Less than",
    RMSLE_invivoPKfit > RMSLE_restrictive &
      RMSLE_invivoPKfit > RMSLE_nonrestrictive~ "More than"
  )) %>% 
  ggplot(aes(x = Chemical, y = RMSLE_invivoPKfit)) +
  geom_point(color = "green4") +
  geom_point(aes(y = RMSLE_nonrestrictive), shape = 6, size = 0.5) +
  geom_point(aes(y = RMSLE_restrictive), shape = 4, size = 0.5) +
  geom_linerange(aes(ymin = RMSLE_nonrestrictive, ymax = RMSLE_restrictive)) +
  facet_grid(~factor(comparison_group,
                     levels = c("Less than", "Between", "More than")),
             scales = "free_x", space = "free_x") +
  labs(y = "RMSLE") +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid = element_blank(),
        axis.title = element_text(size = 14),
        strip.text = element_text(size = 12))

ggsave(filename = paste0(Sys.getenv("FIG_DIR"),
                         "RestrictiveClearance_Analysis/",
                         "CURRENT_", "RMSLE_comparison_fullChart.png"),
       bg = "white",
       height = 3.8, width = 10
)# Number of chemicals we have RMSLE's for
combined_RMSLE %>% pull(Chemical) %>% unique() %>% length()

Another weight that we may consider is relative log-likelihood. This can be
calculated using AIC, compared to the the flat model.

```{r relative-likelihood}

# AIC and relative likelihood
this_rll_probs <- AIC(this_pk) %>%
  filter_targets() %>%
  distinct(Chemical, Species, model, AIC) %>%
  pivot_wider(names_from = model, values_from = AIC) %>%
  rowwise() %>% 
  mutate(minAIC = min(c_across(starts_with("model_")), na.rm = TRUE)) %>% 
  ungroup() %>%
  mutate(across(starts_with("model_"), \(x) {
    signif(exp((minAIC - x)/2) * 100, 4)
  })) %>%
  mutate(across(starts_with("model_"), \(x) {
    ifelse(x < 1E-6, 0, x)
  })) %>%
  inner_join(this_nonflat_winmodel %>% distinct(Chemical, Species, model)) %>%
  ungroup() %>%
  filter(model_1comp >= 99 | model_2comp >= 99,
        model_flat < 1, Chemical %in% target_chems)

this_rll_probs %>%
  count(model_1comp >= 99, model_2comp >= 99)

# LL from pre-calculated httk predictions
restrictive_ll <- all_httk_preds %>% 
  rename(Conc_est = httk_Preds_restrictive,
         Time_trans = Time) %>%
  select(!httk_Preds_nonrestrictive) %>%
  distinct() %>%
  group_by(Chemical, Species) %>%
  nest(.key = "observations") %>%
  inner_join(coef(this_pk) %>% semi_join(this_winmodel) %>% distinct()) %>%
  rowwise() %>%
  mutate(log_likelihood = log_likelihood(
    par = coefs_vector,
    data = observations,
    data_sigma_group = observations$data_sigma_group,
    dose_norm = FALSE,
    log10_trans = FALSE,
    negative = FALSE,
    includes_preds = TRUE,
    detect
  )) %>%
  select(!c(observations, coefs_vector, Time.Units, Time_trans.Units)) %>%
  mutate(model = "model_3comp2_restrictive",
         AIC = -2 * log_likelihood)

nonrestrictive_ll <- all_httk_preds %>% 
  rename(Conc_est = httk_Preds_nonrestrictive,
         Time_trans = Time) %>%
  select(!httk_Preds_restrictive) %>%
  distinct() %>%
  group_by(Chemical, Species) %>%
  nest(.key = "observations") %>%
  inner_join(coef(this_pk) %>% semi_join(this_winmodel) %>% distinct()) %>%
  rowwise() %>%
  mutate(log_likelihood = log_likelihood(
    par = coefs_vector,
    data = observations,
    data_sigma_group = observations$data_sigma_group,
    dose_norm = FALSE,
    log10_trans = FALSE,
    negative = FALSE,
    includes_preds = TRUE,
    detect
  )) %>%
  select(!c(observations, coefs_vector, Time.Units, Time_trans.Units)) %>%
  mutate(model = "model_3comp2_nonrestrictive",
         AIC = -2 * log_likelihood) # Simple formula when there are no optimized params

httk_AIC <- bind_rows(restrictive_ll,
                      nonrestrictive_ll,
                      AIC(this_pk) %>% filter(model == "model_flat") %>%
                        filter_targets())

httk_rel_ll <- httk_AIC %>%
  distinct(Chemical, Species, model, AIC) %>%
  pivot_wider(names_from = model, values_from = AIC) %>%
  rowwise() %>% 
  mutate(minAIC = min(c_across(starts_with("model_")), na.rm = TRUE)) %>% 
  ungroup() %>%
  mutate(across(starts_with("model_"), \(x) {
    signif(exp((minAIC - x)/2) * 100, 4)
  })) %>%
  mutate(across(starts_with("model_"), \(x) {
    ifelse(x < 1E-6, 0, x)
  })) %>%
  mutate(ll_label = factor(
    case_when(model_flat > 95 ~ "Undetermined",
              model_3comp2_nonrestrictive > 95 &
                model_3comp2_restrictive > 95 ~ "Undetermined (non-flat)",
              model_3comp2_nonrestrictive > 95 ~ "Nonrestrictive",
              model_3comp2_restrictive > 95 ~ "Restrictive")
  ))

httk_rel_ll %>%
  count(model_flat > 95, model_3comp2_nonrestrictive > 95,
        model_3comp2_restrictive > 95)

httk_rel_ll %>%
  select(!minAIC) %>%
  mutate(Chemical = forcats::fct_reorder2(Chemical,
                                          model_3comp2_nonrestrictive,
                                          model_flat,
                                          .fun = mean)) %>%
  pivot_longer(cols = starts_with("model_"),
               names_to = "model", values_to = "relative_likelihood") %>%
  ggplot(aes(x = Chemical, y = model)) +
  geom_tile(aes(fill = relative_likelihood)) +
  scale_fill_viridis_b(option = "cividis", breaks = c(0, 50, 80, 100)) +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_text(size = 12, color = "black"),
        axis.ticks = element_blank(),
        panel.grid = element_blank(),
        panel.border = element_blank())

total_rel_evaluation_df <- httk_rel_ll %>%
  inner_join(this_evaluated_df) %>%
  inner_join(this_param_df %>% distinct(Chemical, forced_human_values)) %>%
  distinct()

total_rel_evaluation_df %>%
  distinct(Chemical, Species, Css.tkstats, Css_httk_nr, Css_httk_r, Css_clear,
         ll_label) %>%
  filter(ll_label %in% c("Nonrestrictive", "Restrictive", "Undetermined")) %>%
  ggplot(aes(x = forcats::fct_reorder(Chemical, Css_httk_r))) +
  geom_pointrange(aes(y = Css.tkstats, ymin = Css_httk_nr, ymax = Css_httk_r),
                  size = 0.1) +
  facet_grid(cols = vars(ll_label), space = "free_x", scales = "free_x") +
  scale_y_log10() +
  labs(x = 'Chemicals', y = "Css") +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        strip.text = element_blank())

total_rel_evaluation_df %>%
  distinct(Chemical, Species, Css_httk_r,
           halflife.tkstats, restrictive_halflife,
           nonrestrictive_halflife, halflife_clear,
           ll_label) %>%
    filter(ll_label %in% c("Nonrestrictive", "Restrictive", "Undetermined")) %>%
  ggplot(aes(x = forcats::fct_reorder(Chemical, restrictive_halflife))) +
  geom_pointrange(aes(y = halflife.tkstats, 
                      ymin = nonrestrictive_halflife,
                      ymax = restrictive_halflife),
                  size = 0.1) +
  facet_grid(cols = vars(ll_label), space = "free_x", scales = "free_x") +
  scale_y_log10() +
  labs(x = "Chemicals", y = "halflife") +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        strip.text = element_blank(),
        strip.background = element_blank(),
        axis.text.y = element_text(color = "black", size = 10),
        axis.title = element_text(size = 14))

ggsave(filename = paste0(Sys.getenv("FIG_DIR"),
                         "RestrictiveClearance_Analysis/",
                         "CURRENT_", "halflife_fullChart.png"),
       bg = "white",
       height = 3, width = 12
)
```

The weights to be applied are calculated as such:

$$
\gamma = \frac{100\cdot \rm{RMSLE}}{\theta}
$$
Where $\theta$ is the relative log-likelihood. The smaller number the better.
If the RMSLE is zero, then it means that it is a "perfect" fit and we are highly
confident in using other metrics to gauge clearance. Below I put these together.

>>>>

---
title: "Clearance Project - 4 - PhysChem Clustering"
author: "Gilberto Padilla Mercado"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r data_loading}
target_chem_properties <- readRDS("data-raw/ctxR_3comp2_cheminfo.rds")
new_pk <- readRDS("data-raw/ratCVT_wHTTK_reformulated.rds")
red_assay_res <- readRDS("data-raw/2019-05-06-REDassay_HTTK.rds")
```

@@

target_chem_properties %>%
  rename(Chemical = dtxsid) %>%
  filter_targets() %>%
  distinct(Chemical, monoisotopicMass) %>%
  ggplot(aes(x = monoisotopicMass)) +
  geom_histogram(binwidth = 50, fill = "grey5",
                 boundary = 0) +
  geom_vline(xintercept = 500, linetype = "longdash") +
  scale_x_continuous(limits = c(0, 1000),
                     breaks = seq(0, 1000, 100)) +
  theme_bw()

@@

```{r reticulate-prep}
smiles_list <- target_chem_properties %>% 
  rename(Chemical = dtxsid) %>%
  filter_targets() %>%
  pull(qsarReadySmiles) %>% unique()
mykit <- import("rdkit")
```

@@

```{r chemical_summary_matrix}

target_chem_summary <- target_chem_properties %>%
  filter(propertyId == "logkow-octanol-water", source == "ACD/Labs") %>%
  pivot_wider(names_from = propertyId, values_from = value) %>%
  rename(logkow = `logkow-octanol-water`) %>%
  distinct(dtxsid, monoisotopicMass, qsarReadySmiles, logkow) %>%
  filter(qsarReadySmiles %in% smiles_list) %>%
  right_join(data.frame(qsarReadySmiles = smiles_list,
                        H_donors = py_to_r(py$py_h_donors),
                        H_acceptors = py_to_r(py$py_h_acceptors),
                        TPSA = py_to_r(py$py_tpsa),
                        Rot_Bonds = py_to_r(py$py_rotbonds))
            ) %>%
  rowwise() %>%
  mutate(logD_7.4 = httk::calc_dow(dtxsid = dtxsid, pH = 7.4)) %>%
  left_join(red_assay_res %>%
              distinct(DTXSID, Affinity.Kd.Med) %>%
              na.omit(),
            by = join_by(dtxsid == DTXSID)) %>%
  distinct() %>%
  left_join(cyp_data, join_by(dtxsid == Chemical))

# Make a matrix of all properties so far
tcs_mat <- target_chem_summary %>% 
  select(dtxsid, monoisotopicMass, Affinity.Kd.Med,
         H_donors:Rot_Bonds, logkow, apKa, bpKa, logD_7.4,
         starts_with("CYP")) %>%
  distinct()
tcs_mat <- column_to_rownames(tcs_mat, var = "dtxsid") %>% as.matrix()

# z-score all properties
# Margin = 2 is columns
tcs_mat_z <- apply(tcs_mat, MARGIN = 2, FUN = scale)
rownames(tcs_mat_z) <- rownames(tcs_mat)
summary(tcs_mat_z[, 1])
summary(tcs_mat_z[(rownames(tcs_mat_z) %in% target_chems), 1])
tcs_mat_z[is.na(tcs_mat_z)] <- 0
pca_tcs <- prcomp(tcs_mat_z, scale = FALSE)

pca_df <- pca_tcs$x %>% as.data.frame() %>% rownames_to_column(var = "Chemical")

ggplot(data = pca_df %>% mutate(Target = (Chemical %in% target_chems)) %>%
         left_join(target_chem_summary, by = join_by(Chemical == dtxsid)),
       aes(x = PC1, y = PC2)) +
  geom_point(aes(color = TPSA), alpha = 0.8) +
  scale_color_viridis_c(option = "magma") +
  facet_wrap(~Target) +
  theme_bw()

target_chem_lipinski <- target_chem_summary %>%
  select(-c(qsarReadySmiles)) %>%
  group_by(dtxsid) %>%
  mutate("Mass <= 500 Da" = (monoisotopicMass <= 500),
          "H-donors < 5" = (H_donors < 5),
          "H-acceptors < 10" = (H_acceptors < 10),
          "logP < 5" = (logkow < 5),
          .keep = "none") %>%
  ungroup() %>%
  mutate(Lipinski = rowSums(dplyr::select(., where(is.logical))) >= 3)

lipinski_plot <- target_chem_lipinski %>%
  group_by(Lipinski) %>%
  count(`Mass <= 500 Da`, `H-donors < 5`, `H-acceptors < 10`, `logP < 5`,
        name = "Count") %>%
  ggplot(aes(
    axis3 = `Mass <= 500 Da`,
    axis4 = `logP < 5`,
    axis2 = `H-donors < 5`,
    axis1 = `H-acceptors < 10`,
    y = Count
  )) +
  geom_alluvium(aes(fill = Lipinski)) +
  geom_stratum() +
  scale_x_discrete(limits = c( 
    "H-acceptors < 10",
    "H-donors < 5",
    "<= 500 Da",
    "logP < 5"
  ),
  position = "top",
  expand = c(.2, .05)) +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)),
            size = 1.5) +
  labs(title = "Target chemicals passing Lipinksi's Rules of 5",
       fill = "Passes Lipinksi's Rules?") +
  scale_fill_manual(values = c("grey20", "green4")) +
  theme_minimal() +
  theme(axis.text.x = element_text(face = "bold", color = "black", size = 8),
        panel.grid = element_blank(),
        plot.title = element_text(hjust = 0.5, size = 10),
        axis.title = element_text(size = 8),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 8),
        axis.title.x = element_blank(),
        legend.position = "bottom")

lipinski_plot
ggsave(filename = paste0(Sys.getenv("FIG_DIR"),
                         "RestrictiveClearance_Analysis/",
                         "CURRENT_", "Alluvium_Lipinski.png"),
       bg = "white",
       height = 3, width = 4
)

@@
